### 2014-10-13 doc/thesis/system.tex
Adding more text about subtyping

In the previous chapter we presented an informal overview of Typed Lua.
We showed that programmers can use Typed Lua to combine static and dynamic
typing in the same code, allowing them to incrementally migrate from
dynamic to static typing.
This is a benefit to programmers that use dynamically typed languages
to build large applications, as static types detect many bugs
during the development phase, and also provide better documentation.

In this chapter we present the formalization of Typed Lua.
Besides its practical contribution, Typed Lua also has some interesting
contributions on the field of optional type systems for scripting
languages that we explain in this chapter.
Namely, it is a simple type system that covers several Lua idioms
and features, such as the refinement of tables, multiple return values,
and optional parameters.
We use typing rules to explain how we implemented the features of Typed Lua.

@@

Our type system uses subtyping \citep{abadi1996to,cardelli1984smi} to order
types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We will explain these two relations along this section.
Firt, we will explain the subtyping relationship while we introduce
each type that is present on our type system.
Then, we will explain the differences between subtyping and
consistent-subtyping, showing that they are orthogonal relations.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{f_{i}, ..., f_{n}\}_{c|o|u} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; x_{i} & \textit{projection types}\\
l ::= & \False \; | \; \True \; | \; {\it number} \; | \; {\it string} & \\
b ::= & \Boolean \; | \; \Number \; | \; \String & \\
f ::= & k:t \; | \; \Const \; k:t & \textit{field types}\\ 
k ::= & l \; | \; b \; | \; \Any & \textit{key types}\\
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \Void \; | \; s & \textit{parameters list type}\\
r ::= & \Void \; | \; s \; | \; s \cup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Literal types represent the type of boolean, number, and string
constants.
We shall see that literal types are important to type tables as
records.
The following subtyping rule shows that each literal type
is a subtype of itself:
\[
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash l \subtype l
\end{array}
\]

The variable $\senv$ is a set of pairs of recursion variables that
records the pairs of valid recursive types.

The base types $\Boolean$, $\Number$, and $\String$ represent the
type of values that during run-time Lua tags as booleans, numbers,
and strings, respectively.
Literal types are subtypes of their respective base types, and
each base type is a subtype of itself, as the following subtyping
rules show:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash {\it number} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\;
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash b \subtype b
\end{array}
\end{array}
\]

The type $\Nil$ represents both the type of the constant \texttt{nil}
and the type of values that during run-time Lua tags as nil.
It is a separated type because in our type system $\Nil$ is
not the bottom type.
Typed Lua has this behavior because in Lua the tag \texttt{nil}
represents the absence of an useful value.
Thus, making $\Nil$ the bottom type would severally reduce the
amount of type safety in our type system.
The next subtyping rule shows this behavior:
\[
\begin{array}{c}
\mylabel{S-NIL}\\
\senv \vdash \Nil \subtype \Nil
\end{array}
\]

@@

If we set the type $\Any$ as both bottom and top types of our type
system, then all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down cast any type to $\Any$ and then up cast
$\Any$ to any type.
We shall see later in this section that we use consistent-subtyping
to allow the dynamic type to interact with other first-level types,
and thus allow dynamically typed code coexist with statically typed
code.

Typed Lua uses the type $\Self$ to represent the type of the \texttt{self}
parameter in object-oriented Lua code.
It is only a subtype of itself:
\[
\begin{array}{c}
\mylabel{S-SELF}\\
\senv \vdash \Self \subtype \Self
\end{array}
\]

Union types represent types that can hold a value of several
different, but fixed types.
Notice that union types are disjoint, that is, all types in
the union are not subtypes or supertypes among themselfs.
Next we show the subtyping rules for union types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
       \senv \vdash t_{2} \subtype t}
      {\senv \vdash t_{1} \cup t_{2} \subtype t}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash t \subtype t_{1}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash t \subtype t_{2}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\end{array}
\]

The first rule shows that a union type $t_{1} \cup t_{2}$
is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
of $t$;
while the other rules show that a type $t$ is subtype
of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
$t_{1}$ or if $t$ is subtype of $t_{2}$.

@@

In our type system function types have the form $p \rightarrow r$,
where $p$ is the type of the parameters list and $r$ is the
return type.
The type of the parameters list can be $\Void$ or a tuple type,
while the return type can be $\Void$, a tuple type, or a union of tuples.
The type $\Void$ means an empty tuple.
Our type system includes union of tuples on the return type because
in Lua it is common to overload the return type to denote error,
as we mentioned on Section \ref{sec:statistics}.

@@

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash f_{j} \subtype f_{i}}
      {\senv \vdash \{f_{1}, ..., f_{m}\}_{c} \subtype \{f_{1}, ..., f_{n}\}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash f_{i}' \subtype f_{i}}
      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f_{1}', ..., f_{m}', ..., f_{n}'\}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash f \subtype f_{i}}
      {\senv \vdash \{f_{1}, ..., f_{m}\}_{o} \subtype \{f\}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash t_{2} \subtype t_{1}}
      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

Our type system also includes recursive types: $\mu x.t$,
where $t$ is a function of table type with $x$ appearing anywhere
a function of table type could appear.
For instance, $\mu x.\{``info":\Number, ``next":x \cup \Nil\}$
represents the type of single linked lists of numbers.
We introduce the following subtyping rules for recursive types:

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-VARIABLE}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv, x_{1} \subtype x_{2} \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-AMBERL}\\
\dfrac{\senv \vdash t_{1}[x \mapsto \mu x.t_{1}] \subtype t_{2}}
      {\senv \vdash \mu x.t_{1} \subtype t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBERR}\\
\dfrac{\senv \vdash t_{1} \subtype t_{2}[x \mapsto \mu x.t_{2}]}
      {\senv \vdash t_{1} \subtype \mu x.t_{2}}
\end{array}
\end{array}
\]

Our type system also introduces projection types.
We shall see in the next section that our type system uses
projection types to handle union of typle types when they
appear on the right-hand side of the declaration of local variables.
For this reason, they do not have subtyping rules.

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash t \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim t
\end{array}
\end{array}
\]

@@

Rule for checking if a table type is well formed:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

>>>>

In the previous chapter we presented an informal overview of Typed Lua.
We showed that programmers can use Typed Lua to combine static and dynamic
typing in the same code, and it allows them to incrementally migrate from
dynamic to static typing.
This is a benefit to programmers that use dynamically typed languages
to build large applications, as static types detect many bugs
during the development phase, and also provide better documentation.

In this chapter we present the formalization of Typed Lua.
Besides its practical contribution, Typed Lua also has some interesting
contributions on the field of optional type systems for scripting
languages.
Namely, it is a simple type system that covers several Lua idioms
and features, such as the refinement of tables, multiple return values,
and optional parameters.
We use typing rules to explain how we implemented the features of Typed Lua.

@@

Our type system uses subtyping \citep{cardelli1984smi,abadi1996to} to order
types and consistent-subtyping \citep{siek2007objects,siek2013mutable}
to allow the interaction between statically and dynamically typed code.
We will explain the subtyping and consistent-subtyping rules along this
section while we introduce each type from our type system.
We will focus the discussion on the definition of subtyping because,
as we mentioned in Chapter \ref{chap:review}, we achieve the definition of
consistent-subtyping combining consistency and subtyping.
The differences between subtyping and consistent-subtyping are the way
they handle the dynamic type, and the fact that subtyping is transitive,
but consistent-subtyping is not.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-level types}}\\
t ::= & \;\; l & \textit{literal types}\\
& | \; b & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; t \cup t & \textit{disjoint union types}\\
& | \; p \rightarrow r & \textit{function types}\\
& | \; \{v_{i}, ..., v_{n}\}_{closed|open|unique} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.t & \textit{recursive types}\\
& | \; \pi_{i} & \textit{projection types}\\
l ::= & \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} & \\
b ::= & \Boolean \; | \; \Integer \; | \; \Number \; | \; \String & \\
v ::= & k:t \; | \; \Const \; k:t & \textit{field types}\\ 
k ::= & l \; | \; b \; | \; \Any & \textit{key types}\\
\multicolumn{3}{c}{}\\
\multicolumn{3}{c}{\textbf{Second-level types}}\\
s ::= & \;\; t &\\
& | \; t* & \textit{variadic types}\\
& | \; t \times s & \textit{tuple types}\\
p ::= & \Void \; | \; s & \textit{parameters list type}\\
r ::= & \Void \; | \; s \; | \; s \sqcup r & \textit{return types}\\
\end{array}
$$
\dend
\caption{The abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

@@

Literal types represent the type of literal constants.
They can be the boolean values $\False$ and $\True$,
any integer value, any floating point value, and any string value.
We shall see that literal types are important to type tables as records.

Our type system also includes the following base types: $\Boolean$,
$\Integer$, $\Number$, and $\String$.
The base types $\Boolean$ and $\String$ represent the values that
during run-time Lua tags as \texttt{boolean} and \texttt{string},
respectively;
while the base types $\Integer$ and $\Number$ represent the values
that during run-time Lua tags as \texttt{number}.
We introduced two different base types to represent Lua numbers
because Lua introduced integers recently.

The subtyping rules for literal types include a reflexive rule
and the rules for defining that literal types are subtypes of
their respective base types;
the subtyping rules for base types include a reflexive rule
and another rule for defining that $\Integer$ is subtype of
$\Number$.
We defined these subtyping rules as follows:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-LITERAL}\\
\senv \vdash l \subtype l
\end{array}
\;
\begin{array}{c}
\mylabel{S-FALSE}\\
\senv \vdash \False \subtype \Boolean
\end{array}
\;
\begin{array}{c}
\mylabel{S-TRUE}\\
\senv \vdash \True \subtype \Boolean
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-INT1}\\
\senv \vdash {\it int} \subtype \Integer
\end{array}
\;
\begin{array}{c}
\mylabel{S-INT2}\\
\senv \vdash {\it int} \subtype \Number
\end{array}
\;
\begin{array}{c}
\mylabel{S-FLOAT}\\
\senv \vdash {\it float} \subtype \Number
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-STRING}\\
\senv \vdash {\it string} \subtype \String
\end{array}
\begin{array}{c}
\mylabel{S-BASE}\\
\senv \vdash b \subtype b
\end{array}
\;
\begin{array}{c}
\mylabel{S-NUMBER}\\
\senv \vdash \Integer \subtype \Number
\end{array}
\end{array}
\]

The variable $\senv$ is a set of pairs of recursion variables.
We shall notice that we need this set to record the pairs of valid
recursive types.

Since subtyping is reflexive and transitive,
we could have ommited the reflexive rules \textsc{S-LITERAL} and
\textsc{S-BASE}, as well as the rule \textsc{S-INT2}.
More precisely, we could have used subsumption rules that define
the reflexiveness and transitiveness of subtyping.
In the case of \textsc{S-INT2}, we would not need this rule because
we could use the transitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

However, we are using the subtyping rules as the template for defining
the consistent-subtyping rules, and consistent-subtyping is not
transitive.
More precisely, we want the subtyping and consistent-subtyping rules
to differ only in the way they handle the dynamic type.
Thus, we define the subtyping rules using an algorithmic
way that is close to the implementation, as this way allows us to use
subtyping to formalize consistent-subtyping.

The type $\Nil$ represents the type of values that during run-time
Lua tags as \texttt{nil}.
It is a separated type because in our type system $\Nil$ is
not the bottom type.
Typed Lua has this behavior because in Lua the tag \texttt{nil}
represents the absence of an useful value.
Thus, making $\Nil$ the bottom type would severally reduce the
amount of type safety in our type system.
The next subtyping rule shows this behavior:
\[
\begin{array}{c}
\mylabel{S-NIL}\\
\senv \vdash \Nil \subtype \Nil
\end{array}
\]

@@

Even though the dynamic type $\Any$ is neutral to subtyping,
it is not to consistent-subtyping:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{C-ANY1}\\
\senv \vdash t \lesssim \Any
\end{array}
\;
\begin{array}{c}
\mylabel{C-ANY2}\\
\senv \vdash \Any \lesssim t
\end{array}
\end{array}
\]

The rules that handle the dynamic type are the only rules that
differ between subtyping and consistent-subtyping.
If we set the type $\Any$ as both bottom and top types of our
subtyping relation, then all programs would type check without errors.
This would happen due to the transitivity of subtyping, that is,
we would be able to down-cast any type to $\Any$ and then up-cast
$\Any$ to any type.
While the rule \textsc{S-ANY} is neutral and makes transitivity
not be a problem, the rules \textsc{C-ANY1} and \textsc{C-ANY2}
are the rules that allow the dynamic type to interact with other
first-level types, and thus allow dynamically typed code coexist
with statically typed code.
Consistent-subtyping cannot be transitive because of these rules,
as transitivity combined with them would make all programs compile
without type errors.

Typed Lua uses the type $\Self$ to represent \emph{objects}.
As we mentioned in Chapter \ref{chap:typedlua}, we need the type
$\Self$ to prevent programmers from indexing a method without
calling it with the correct receiver.
The type $\Self$ is only a subtype of itself:
\[
\begin{array}{c}
\mylabel{S-SELF}\\
\senv \vdash \Self \subtype \Self
\end{array}
\]

Union types represent types that can hold a value of several
different, but fixed types.
Next we show the subtyping rules for union types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-UNION1}\\
\dfrac{\senv \vdash t_{1} \subtype t \;\;\;
       \senv \vdash t_{2} \subtype t}
      {\senv \vdash t_{1} \cup t_{2} \subtype t}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION2}\\
\dfrac{\senv \vdash t \subtype t_{1}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UNION3}\\
\dfrac{\senv \vdash t \subtype t_{2}}
      {\senv \vdash t \subtype t_{1} \cup t_{2}}
\end{array}
\end{array}
\]

The first rule shows that a union type $t_{1} \cup t_{2}$
is subtype of $t$ if both $t_{1}$ and $t_{2}$ are subtypes
of $t$;
and the other rules show that a type $t$ is subtype
of a union type $t_{1} \cup t_{2}$ if $t$ is subtype of
$t_{1}$ or if $t$ is subtype of $t_{2}$.

Union types are disjoint, that is, the types in the union are not
consistent-subtype of any other type in the same union.
For instance, the union type $\Boolean \;\cup\; \Any$ results the
type $\Any$, because $\Boolean$ is consistent-subtype of $\Any$.
The union type $\Number \;\cup\; \Nil \;\cup\; 1$ results the union
type $\Number \;\cup\; \Nil$ because $1 \subtype \Number$.

@@

In our type system function types have the form $p \rightarrow r$,
where $p$ is the type of the parameters list and $r$ is the
return type.
The type of the parameters list can be $\Void$ or a tuple type,
while the return type can be $\Void$, a tuple type, or a union of tuples.
The type $\Void$ means an empty tuple.
Our type system includes union of tuples on the return type because
in Lua it is common to overload the return type to denote error,
as we mentioned in Section \ref{sec:statistics}.

@@

Table types have only one syntactical form, but they also have one of
these three tags: \emph{closed}, \emph{open}, and \emph{unique}.
The tag \emph{closed} is the tag of the table types that the type
annotations describe.
The tag \emph{open} is the tag of the table types that we can
add new fields through field assignment.
The tag \emph{unique} is the tag of the table types that represent
the type of the table constructors.
The syntactical form of table types is $\{ v_{i}, ..., v_{n} \}$,
where $i$ varies from 1 to $n$ when $n > 0$,
but $i$ can also be 0 when $n = 0$.

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE1}\\
\dfrac{\forall i \in 1..n \; \exists j \in 1..m \;\;\;
       \senv \vdash v_{j} \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{closed} \subtype \{v_{1}, ..., v_{n}\}_{closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{open} \subtype \{v\}_{open|closed}} 
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v_{i}' \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v_{1}', ..., v_{m}', ..., v_{n}'\}_{unique|open|closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\forall i \in 1..m \; \senv \vdash v \subtype_{n} v_{i}}
      {\senv \vdash \{v_{1}, ..., v_{m}\}_{unique} \subtype \{v\}_{unique|open|closed}} 
\end{array}
\end{array}
\]

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2} \;\;\;
       \senv \vdash t_{2} \subtype t_{1}}
      {\senv \vdash k_{1}:t_{1} \subtype k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD4}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} k_{2}:t_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD5}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \Const \; k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD6}\\
\dfrac{\senv \vdash k_{2} \subtype k_{1} \;\;\;
       \senv \vdash t_{1} \subtype t_{2}}
      {\senv \vdash k_{1}:t_{1} \subtype_{n} \Const \; k_{2}:t_{2}}
\end{array}
\end{array}
\]

Table types can be ambiguous.
For instance, $\{1:\Number, \Number:\String\}$ is an ambiguous
table type because the type of the value stored by key $1$ can be
$\Number$ or $\String$, as $1 \subtype 1$ and
$1 \subtype \Number$.
Table types can get even more ambiguous when we use the dynamic
type in the type of table keys.
For instance, the table type $\{1:\Number, \Number:\String, \Any:\Boolean\}$
is ambiguous because the type of the value stored by key $1$ can be
$\Number$, or $\String$, or $\Boolean$, as $1 \subtype 1$,
$1 \subtype \Number$, and $1 \lesssim \Any$.
The type of the value stored by a key of type $\Number$,
which is not the literal type $1$, can be $\Number$ or $\Boolean$,
as $\Number \subtype \Number$, and $\Number \lesssim \Any$.
To avoid ambiguous table types we use the following rule to check
whether a table type is well formed:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

Our type system also includes recursive types: $\mu x.t$,
where $t$ is a function or table type with $x$ appearing anywhere
a function or table type could appear.
For instance, $\mu x.\{``info":\Number, ``next":x \;\cup\; \Nil\}$
represents the type of single linked lists of numbers.

We use the \emph{Amber rule} from the Amber language
\citep{cardelli1986amber} to define the subtyping rule for
recursive types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-ASSUMPTION}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
      {\senv \vdash x_{1} \subtype x_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-AMBER}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash t_{1} \subtype t_{2}}
      {\senv \vdash \mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}}
\end{array}
\end{array}
\]

The rule \textsc{S-AMBER} needs the rule \textsc{S-ASSUMPTION}
to show that $\mu x_{1}.t_{1} \subtype \mu x_{2}.t_{2}$.
The rule \textsc{S-ASSUMPTION} checks if the assumption
$x_{1} \subtype x_{2}$ exists in the set of assumptions $\senv$.
The rule \textsc{S-AMBER} extends the set of assumptions
with the assumption $x_{1} \subtype x_{2}$ to check whether
$t_{1} \subtype t_{2}$ or not.

Our type system also introduces projection types, but they do not
appear on the subtyping and consistent-subtyping rules.
They are just a mechanism that the type system uses for handling
union of tuple types, when they appear on the right-hand side of
the declaration of local variables.
That is the reason why we do not need to define subtyping and
consistent-subtyping rules for them.
We will see projection types in more detail in the next section.

@@
